---
title: 'GBIF Small Mammal Occurrence Analysis for Central Asia: Kazakhstan Field Guide'
author: "Central Asia One Health Project - ZSL"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: '3'
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: kate
    code_folding: show
    fig_width: 10
    fig_height: 8
subtitle: A comprehensive protocol for small mammal surveillance planning
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = 'center',
  fig.width = 10,
  fig.height = 8
)

# Set root directory for all chunks
knitr::opts_knit$set(root.dir = "~/GitHub/Rodent_CA/")

# Load all required libraries
library(tidyverse)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(viridis)
library(scales)
library(patchwork)
library(lubridate)
library(cowplot)
library(kableExtra)
library(plotly)
library(DT)

# Set consistent theme
theme_set(theme_minimal(base_size = 12))

# Define consistent colour palettes
species_colors <- colorRampPalette(c("#440154", "#31688e", "#35b779", "#fde725"))(20)
month_colors <- c("#2E7D32", "#43A047", "#66BB6A", "#81C784", 
                  "#FDD835", "#FFEB3B", "#FFF176", "#FFB74D", 
                  "#FF8A65", "#A1887F", "#8D6E63", "#6D4C41")
family_colors <- c("#E53935", "#D81B60", "#8E24AA", "#5E35B1", 
                   "#3949AB", "#1E88E5", "#039BE5", "#00ACC1", 
                   "#00897B", "#43A047", "#7CB342", "#C0CA33")
```

```{r params}
# Analysis parameters
# TO MODIFY: Change these to analyse a different dataset or focal country
# The previous file ("1_Data_Sets/01_GBIF/Rodents_kazakhstan.csv", Rodentia and
# Kazakhstan only) still works if substituted here
data_file <- "1_Data_Sets/01_GBIF/Small_mammals_CA.csv"
focal_iso <- "KZ"
focal_name <- "Kazakhstan"
```

# Introduction

This document provides a comprehensive analysis of small mammal occurrence data from Kazakhstan using GBIF (Global Biodiversity Information Facility) records. It is designed to help field teams prepare for small mammal surveillance activities by understanding:

- **What species** you might encounter
- **Where** different species have been recorded
- **When** small mammals are most likely to be captured
- **How reliable** the existing data is

> **Note**: This analysis is based on opportunistic occurrence records from GBIF. Patterns may reflect sampling effort and researcher activity rather than true biological patterns. 

# Data Loading and Initial Exploration

## Load the GBIF Data
```{r load-data}
# Read the GBIF data
# TO MODIFY: Change data_file in the params chunk to your data file location
gbif_data_raw <- read.csv(data_file,
                          sep = "\t",
                          stringsAsFactors = FALSE)

# Display basic information
cat("Dataset Overview:\n")
cat("Total records:", format(nrow(gbif_data_raw), big.mark = ","), "\n")
cat("Number of variables:", ncol(gbif_data_raw), "\n")
cat("Date range:", min(gbif_data_raw$year, na.rm = TRUE), "-", 
    max(gbif_data_raw$year, na.rm = TRUE), "\n\n")
```

## Data Cleaning and Preparation
```{r clean-data}
# How many records have coordinates?
cat("Records WITH coordinates:", sum(!is.na(gbif_data_raw$decimalLatitude) & 
                                      !is.na(gbif_data_raw$decimalLongitude)), "\n")

cat("Records WITHOUT coordinates:", sum(is.na(gbif_data_raw$decimalLatitude) | 
                                         is.na(gbif_data_raw$decimalLongitude)), "\n")

# Are there any invalid coordinates?
cat("Invalid latitudes:", sum(gbif_data_raw$decimalLatitude < -90 | 
                               gbif_data_raw$decimalLatitude > 90, na.rm = TRUE), "\n")

cat("Invalid longitudes:", sum(gbif_data_raw$decimalLongitude < -180 | 
                                gbif_data_raw$decimalLongitude > 180, na.rm = TRUE), "\n")

# What's in the original species column?
table(is.na(gbif_data_raw$species) | gbif_data_raw$species == "")

# What's in genus column?
table(is.na(gbif_data_raw$genus) | gbif_data_raw$genus == "")


## Data Cleaning and Preparation

# Report how many records the fossil exclusion removes
n_fossil <- sum(gbif_data_raw$basisOfRecord == "FOSSIL_SPECIMEN" |
                  gbif_data_raw$datasetKey == "bb5b30b4-827e-4d5e-a86a-825d65cb6583",
                na.rm = TRUE)
cat("Palaeontological records excluded (PBDB dataset or FOSSIL_SPECIMEN):", n_fossil, "\n")

gbif_nofossil <- gbif_data_raw %>%
  # Exclude palaeontological records: the Paleobiology Database publishes
  # fossil occurrences mostly as MATERIAL_CITATION, so basisOfRecord alone
  # does not catch them
  filter(basisOfRecord != "FOSSIL_SPECIMEN",
         datasetKey != "bb5b30b4-827e-4d5e-a86a-825d65cb6583") %>%
  # Add cleaned/standardised variables (keep originals)
  mutate(
    # Standardise species names
    species_clean = case_when(
      !is.na(species) & species != "" ~ species,
      !is.na(genus) & genus != "" ~ paste(genus, "sp."),
      TRUE ~ scientificName
    ),
    # Standardise family names
    family_clean = if_else(
      is.na(family) | family == "",
      "Unknown",
      family
    ),
    # Standardise order names
    order_clean = if_else(is.na(order) | order == "", "Unknown", order),
    # Ensure dates are correct format
    year = as.integer(year),
    month = as.integer(month),
    day = as.integer(day),
    # Categorise coordinate precision
    coord_precision = case_when(
      coordinateUncertaintyInMeters <= 100 ~ "High (<100m)",
      coordinateUncertaintyInMeters <= 1000 ~ "Medium (100-1000m)",
      coordinateUncertaintyInMeters <= 10000 ~ "Low (1-10km)",
      coordinateUncertaintyInMeters > 10000 ~ "Very Low (>10km)",
      TRUE ~ "Unknown"
    )
  )

# Mapping dataset: keep georeferenced records only. The species inventory is
# built from gbif_nofossil, so records without coordinates still count there
gbif_data <- gbif_nofossil %>%
  # Remove records without coordinates
  filter(!is.na(decimalLatitude) & !is.na(decimalLongitude)) %>%
  # Remove invalid coordinates
  filter(
    decimalLatitude >= -90 & decimalLatitude <= 90,
    decimalLongitude >= -180 & decimalLongitude <= 180
  )

# Show first 20 species_clean values to check logic
head(gbif_data$species_clean, 20)

# What's in original family column?
table(is.na(gbif_data_raw$family) | gbif_data_raw$family == "")

# Show unique families
unique(gbif_data$family_clean)
# Print summary
cat("Data Cleaning Summary:\n")
cat("Non-fossil records:", format(nrow(gbif_nofossil), big.mark = ","), "\n")
cat("Georeferenced records for mapping:", format(nrow(gbif_data), big.mark = ","),
    "(", round(nrow(gbif_data) / nrow(gbif_nofossil) * 100, 1), "% retained)\n")
cat("Unique species:", n_distinct(gbif_data$species_clean), "\n")
cat("Unique families:", n_distinct(gbif_data$family_clean), "\n")
cat("Unique genera:", n_distinct(gbif_data$genus), "\n")


## Split into two levels: whole Central Asia (reference) and focal country

# Keep the full cleaned Central Asia dataset as a reference layer
gbif_ca <- gbif_data

# Restrict gbif_data to the focal country so all downstream chunks
# analyse the focal country only
gbif_data <- gbif_ca %>%
  filter(countryCode == focal_iso)

# Compare record counts: total Central Asia vs focal country, per order
cat("\nRecord Counts: Central Asia vs", focal_name, "\n")
cat("Total Central Asia records:", format(nrow(gbif_ca), big.mark = ","), "\n")
cat(focal_name, "records:", format(nrow(gbif_data), big.mark = ","), "\n\n")

order_comparison <- gbif_ca %>%
  group_by(order_clean) %>%
  summarise(
    n_ca = n(),
    n_focal = sum(countryCode == focal_iso, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  arrange(desc(n_ca))

print(order_comparison)
```

# Species Diversity Analysis

## Complete Species Inventory

```{r species-inventory}
# Create comprehensive species list from all non-fossil records: a museum
# specimen without coordinates is still evidence a species occurs
species_summary <- gbif_nofossil %>%
  filter(countryCode == focal_iso) %>%
  mutate(
    is_georef = !is.na(decimalLatitude) & !is.na(decimalLongitude) &
      decimalLatitude >= -90 & decimalLatitude <= 90 &
      decimalLongitude >= -180 & decimalLongitude <= 180
  ) %>%
  group_by(species_clean, order_clean, family_clean) %>%
  summarise(
    n_records = n(),
    n_georeferenced = sum(is_georef),
    n_locations = n_distinct(paste(round(decimalLatitude, 2),
                                   round(decimalLongitude, 2))[is_georef]),
    year_range = paste(min(year, na.rm = TRUE), "-",
                      max(year, na.rm = TRUE)),
    latest_year = if (all(is.na(year))) NA_integer_ else max(year, na.rm = TRUE),
    n_provinces = n_distinct(stateProvince, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  arrange(desc(n_records)) %>%
  mutate(
    percent_records = round(n_records / sum(n_records) * 100, 2),
    cumulative_percent = round(cumsum(n_records) / sum(n_records) * 100, 1),
    note = if_else(n_georeferenced == 0,
                   "collection records only, none georeferenced", "")
  )

# Duplicate species rows would corrupt the common/rare species selection downstream
stopifnot(!any(duplicated(species_summary$species_clean)))

if (knitr::is_html_output()) {
  datatable(species_summary,
            options = list(pageLength = 15, scrollX = TRUE),
            caption = "Complete inventory of small mammal species")
} else {
  kable(head(species_summary, 10) %>% 
          select(species_clean, family_clean, n_records, n_locations, latest_year),
        caption = "Top 10 most common small mammal species",
        col.names = c("Species", "Family", "Records", "Locations", "Latest year"),
        booktabs = TRUE) %>%
    kable_styling(latex_options = c("striped", "HOLD_position"),
                  font_size = 10) %>%
    column_spec(1, italic = TRUE)
}

# Export species inventories for the field guide
dir.create("4_Tables", showWarnings = FALSE)

# Focal-country inventory (expected species list for the trapping trip)
write.csv(species_summary, paste0("4_Tables/species_inventory_", focal_iso, ".csv"),
          row.names = FALSE)

# Whole Central Asia inventory (reference layer for cross-border species)
species_summary_ca <- gbif_nofossil %>%
  mutate(
    is_georef = !is.na(decimalLatitude) & !is.na(decimalLongitude) &
      decimalLatitude >= -90 & decimalLatitude <= 90 &
      decimalLongitude >= -180 & decimalLongitude <= 180
  ) %>%
  group_by(species_clean, order_clean, family_clean) %>%
  summarise(
    n_records = n(),
    n_georeferenced = sum(is_georef),
    n_locations = n_distinct(paste(round(decimalLatitude, 2),
                                   round(decimalLongitude, 2))[is_georef]),
    year_range = paste(min(year, na.rm = TRUE), "-",
                      max(year, na.rm = TRUE)),
    latest_year = if (all(is.na(year))) NA_integer_ else max(year, na.rm = TRUE),
    n_provinces = n_distinct(stateProvince, na.rm = TRUE),
    countries = paste(sort(unique(countryCode)), collapse = ", "),
    in_focal = any(countryCode == focal_iso, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  arrange(desc(n_records)) %>%
  mutate(
    percent_records = round(n_records / sum(n_records) * 100, 2),
    cumulative_percent = round(cumsum(n_records) / sum(n_records) * 100, 1),
    note = if_else(n_georeferenced == 0,
                   "collection records only, none georeferenced", "")
  )

write.csv(species_summary_ca, "4_Tables/species_inventory_CA.csv", row.names = FALSE)
```

## Species Accumulation Curve

```{r species-accumulation, fig.height=6}
# Calculate species accumulation by year
species_accumulation <- gbif_data %>%
  filter(!is.na(year)) %>%
  arrange(year) %>%
  group_by(year) %>%
  summarise(
    new_species = list(unique(species_clean)),
    .groups = 'drop'
  ) %>%
  mutate(
    cumulative_species = accumulate(new_species, union),
    n_cumulative = map_int(cumulative_species, length),
    n_new = map_int(new_species, length)
  )

# Plot accumulation curve
p1 <- ggplot(species_accumulation, aes(x = year)) +
  geom_line(aes(y = n_cumulative), size = 1.5, color = "#2E7D32") +
  geom_point(aes(y = n_cumulative), size = 2, color = "#2E7D32") +
  geom_bar(aes(y = n_new * 3), stat = "identity", alpha = 0.3, fill = "#81C784") +
  scale_y_continuous(
    name = "Cumulative number of species",
    sec.axis = sec_axis(~./3, name = "New species per year")
  ) +
  labs(
    title = "Species Discovery Timeline",
    subtitle = "Cumulative species recorded (line) and new species per year (bars)",
    x = "Year"
  ) +
  theme_minimal() +
  theme(
    axis.title.y.right = element_text(color = "#81C784"),
    plot.title = element_text(size = 14, face = "bold")
  )

print(p1)
```

## Family Composition

```{r family-composition, fig.height=8}
# Prepare family data
family_summary <- gbif_data %>%
  group_by(family_clean) %>%
  summarise(
    n_records = n(),
    n_species = n_distinct(species_clean),
    n_genera = n_distinct(genus),
    .groups = 'drop'
  ) %>%
  arrange(desc(n_records)) %>%
  mutate(
    family_clean = factor(family_clean, levels = family_clean),
    percent = n_records / sum(n_records) * 100
  )

# Create two-panel plot
p1 <- ggplot(family_summary, aes(x = family_clean, y = n_records, fill = family_clean)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = family_colors[1:nrow(family_summary)]) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "A. Number of Records per Family",
    x = "", y = "Number of records"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none",
    plot.title = element_text(face = "bold")
  )

p2 <- ggplot(family_summary, aes(x = family_clean, y = n_species, fill = family_clean)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = family_colors[1:nrow(family_summary)]) +
  labs(
    title = "B. Species Diversity per Family",
    x = "", y = "Number of species"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none",
    plot.title = element_text(face = "bold")
  )

# Combine plots
family_plot <- p1 / p2
print(family_plot)

# Summary statistics
cat("\nFamily Diversity Summary:\n")
cat("Most recorded family:", family_summary$family_clean[1], 
    "with", family_summary$n_records[1], "records\n")
cat("Most species-rich family:", 
    family_summary$family_clean[which.max(family_summary$n_species)],
    "with", max(family_summary$n_species), "species\n")
```

# Spatial Distribution Patterns

## Geographic Coverage Map
```{r coverage-map, fig.height=8}
# Get Kazakhstan boundaries
kazakhstan <- ne_countries(country = "Kazakhstan", scale = "medium", returnclass = "sf")

# Create coverage map with all points
p_coverage <- ggplot() +
  geom_sf(data = kazakhstan, fill = "grey95", color = "grey60") +
  geom_point(data = gbif_data,
             aes(x = decimalLongitude, y = decimalLatitude),
             alpha = 0.5, size = 0.8, color = "red") +
  labs(
    title = "Geographic Coverage of Small Mammal Records in Kazakhstan",
    subtitle = paste("Total:", format(nrow(gbif_data), big.mark = ","), "georeferenced records"),
    x = "Longitude", y = "Latitude"
  ) +
  theme_minimal() +
  coord_sf(xlim = c(46, 88), ylim = c(40, 56))

print(p_coverage)


# Save as PDF instead
ggsave(
  filename = "3_Figures/1_Gbif_overview/Geographic_coverage_kz.pdf",  # Changed to .pdf
  plot = p_coverage, 
  width = 10,
  height = 8,
  units = "in",
  device = "pdf"  # Explicitly specify PDF
)
```

## Regional Coverage Map (Central Asia)
```{r coverage-map-ca, fig.height=8}
# Get boundaries for the five Central Asian countries
central_asia <- ne_countries(country = c("Kazakhstan", "Kyrgyzstan", "Tajikistan",
                                         "Turkmenistan", "Uzbekistan"),
                             scale = "medium", returnclass = "sf")

# Create regional coverage map with all points, coloured by order
p_coverage_ca <- ggplot() +
  geom_sf(data = central_asia, fill = "grey95", color = "grey60") +
  geom_point(data = gbif_ca,
             aes(x = decimalLongitude, y = decimalLatitude, color = order_clean),
             alpha = 0.5, size = 0.8) +
  labs(
    title = "Geographic Coverage of Small Mammal Records in Central Asia",
    subtitle = paste("Total:", format(nrow(gbif_ca), big.mark = ","), "georeferenced records"),
    x = "Longitude", y = "Latitude",
    color = "Order"
  ) +
  theme_minimal() +
  coord_sf(xlim = c(46, 88), ylim = c(35, 56))

print(p_coverage_ca)


ggsave(
  filename = "3_Figures/1_Gbif_overview/Geographic_coverage_CA.pdf",
  plot = p_coverage_ca,
  width = 10,
  height = 8,
  units = "in",
  device = "pdf"  # Explicitly specify PDF
)
```

## Species Richness Hotspots

Where is small mammal diversity highest? 
```{r richness-map, fig.height=8}
# Create grid for species richness
grid_size <- 0.5  # degrees

richness_grid <- gbif_data %>%
  mutate(
    lat_grid = floor(decimalLatitude / grid_size) * grid_size + grid_size/2,
    lon_grid = floor(decimalLongitude / grid_size) * grid_size + grid_size/2
  ) %>%
  group_by(lat_grid, lon_grid) %>%
  summarise(
    species_richness = n_distinct(species_clean),
    n_records = n(),
    n_families = n_distinct(family_clean),
    .groups = 'drop'
  )

# Plot species richness
p_richness <- ggplot() +
  geom_sf(data = kazakhstan, fill = "grey95", color = "grey60") +
  geom_tile(data = richness_grid,
            aes(x = lon_grid, y = lat_grid, fill = species_richness),
            alpha = 0.8) +
  scale_fill_viridis(
    name = "Species\nrichness",
    option = "plasma",
    breaks = pretty_breaks(n = 5)
  ) +
  labs(
    title = "Small Mammal Species Richness Hotspots",
    subtitle = paste("Grid size:", grid_size, "degrees"),
    x = "Longitude", y = "Latitude"
  ) +
  theme_minimal() +
  coord_sf(xlim = c(46, 88), ylim = c(40, 56))

print(p_richness)

# Identify top diversity hotspots
top_hotspots <- richness_grid %>%
  arrange(desc(species_richness)) %>%
  head(10) %>%
  mutate(
    location = paste0("(", round(lat_grid, 1), "°N, ", round(lon_grid, 1), "°E)")
  )

cat("\nTop 10 Diversity Hotspots:\n")
for(i in 1:nrow(top_hotspots)) {
  cat(i, ". ", top_hotspots$location[i], ": ", 
      top_hotspots$species_richness[i], " species (", 
      top_hotspots$n_records[i], " records)\n", sep = "")
}

# Save as PDF instead
ggsave(
  filename = "3_Figures/1_Gbif_overview/Species_richness_kz.pdf",  # Changed to .pdf
  plot = p_richness, 
  width = 10,
  height = 8,
  units = "in",
  device = "pdf"  # Explicitly specify PDF
)
```

## Individual Species Distribution Maps

Interactive maps for the most common species AND rare species
```{r species-maps, fig.height=12}
# Select species to map: top 6 common + 3 rare species with >1 georeferenced
# record (species without georeferenced records cannot be mapped)
common_species <- species_summary %>%
  filter(n_georeferenced > 1) %>%
  head(6) %>%
  pull(species_clean)

rare_species <- species_summary %>%
  filter(n_georeferenced > 1) %>%
  tail(3) %>%
  pull(species_clean)

species_to_map <- c(common_species, rare_species)

# Create individual maps
species_maps <- map(species_to_map, function(sp) {
  sp_data <- gbif_data %>% filter(species_clean == sp)
  
  p <- ggplot() +
    geom_sf(data = kazakhstan, fill = "grey95", color = "grey60") +
    geom_point(data = sp_data,
               aes(x = decimalLongitude, y = decimalLatitude),
               color = "#E53935", alpha = 0.7, size = 1.5) +
    labs(
      title = sp,
      subtitle = paste("n =", nrow(sp_data), "records")
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(face = "italic", size = 10),
      plot.subtitle = element_text(size = 8)
    ) +
    coord_sf(xlim = c(46, 88), ylim = c(40, 56))
  
  return(p)
})

# Arrange in grid
species_grid <- wrap_plots(species_maps, ncol = 3) +
  plot_annotation(
    title = "Distribution Maps: Common Species (top 2 rows) and Rare Species (bottom row)",
    theme = theme(plot.title = element_text(size = 14, face = "bold"))
  )

print(species_grid)



```

# Temporal Patterns

## Seasonal Activity Patterns

**Critical Note**: Seasonal patterns may reflect researcher activity (holidays, field seasons) rather than true small mammal activity
```{r seasonal-patterns, fig.height=10}
# Prepare monthly data
monthly_data <- gbif_data %>%
  filter(!is.na(month)) %>%
  group_by(month) %>%
  summarise(
    n_records = n(),
    n_species = n_distinct(species_clean),
    .groups = 'drop'
  ) %>%
  mutate(
    month_name = month.abb[month],
    month_name = factor(month_name, levels = month.abb)
  )

# Overall monthly pattern
p1 <- ggplot(monthly_data, aes(x = month_name, y = n_records)) +
  geom_bar(stat = "identity", fill = month_colors) +
  geom_line(aes(group = 1, y = n_species * max(n_records) / max(n_species)), 
            color = "red", size = 1.5) +
  scale_y_continuous(
    name = "Number of records",
    labels = comma,
    sec.axis = sec_axis(~. * max(monthly_data$n_species) / max(monthly_data$n_records), 
                        name = "Number of species (red line)")
  ) +
  labs(
    title = "A. Overall Seasonal Pattern",
    subtitle = "⚠️ May reflect sampling effort rather than biological patterns",
    x = "Month"
  ) +
  theme_minimal() +
  theme(
    axis.title.y.right = element_text(color = "red"),
    plot.subtitle = element_text(color = "darkorange", face = "italic")
  )

# Family-specific patterns
family_monthly <- gbif_data %>%
  filter(!is.na(month), family_clean %in% head(family_summary$family_clean, 6)) %>%
  group_by(family_clean, month) %>%
  summarise(n_records = n(), .groups = 'drop') %>%
  mutate(month_name = factor(month.abb[month], levels = month.abb))

p2 <- ggplot(family_monthly, aes(x = month_name, y = n_records, fill = family_clean)) +
  geom_bar(stat = "identity") +
  facet_wrap(~family_clean, scales = "free_y", ncol = 2) +
  scale_fill_manual(values = family_colors[1:6]) +
  labs(
    title = "B. Seasonal Patterns by Family",
    x = "Month", y = "Number of records"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8)
  )

seasonal_plot <- p1 / p2 + plot_layout(heights = c(1, 2))
print(seasonal_plot)


# Save as PDF instead
ggsave(
  filename = "3_Figures/1_Gbif_overview/Seasonal_plot_kz.pdf",  # Changed to .pdf
  plot = seasonal_plot, 
  width = 10,
  height = 8,
  units = "in",
  device = "pdf"  # Explicitly specify PDF
)

# Statistical summary
cat("\nSeasonal Sampling Summary:\n")
cat("Peak collection month:", monthly_data$month_name[which.max(monthly_data$n_records)], 
    "with", max(monthly_data$n_records), "records\n")
cat("Lowest collection month:", monthly_data$month_name[which.min(monthly_data$n_records)], 
    "with", min(monthly_data$n_records), "records\n")
cat("Months with no records:", paste(month.abb[!(1:12 %in% monthly_data$month)], collapse = ", "), "\n")
```

## Long-term Temporal Trends
```{r temporal-trends, fig.height=8}
# Yearly patterns
yearly_data <- gbif_data %>%
  filter(!is.na(year), year >= 1950) %>%  # Focus on modern records
  group_by(year) %>%
  summarise(
    n_records = n(),
    n_species = n_distinct(species_clean),
    n_collectors = n_distinct(recordedBy, na.rm = TRUE),
    .groups = 'drop'
  )

p1 <- ggplot(yearly_data, aes(x = year)) +
  geom_bar(aes(y = n_records), stat = "identity", fill = "#2E7D32", alpha = 0.7) +
  geom_line(aes(y = n_species * 10), color = "#D84315", size = 1.2) +
  scale_y_continuous(
    name = "Number of records",
    labels = comma,
    sec.axis = sec_axis(~./10, name = "Number of species")
  ) +
  labs(
    title = "Collection Effort Over Time",
    subtitle = "Bars: records per year | Line: species diversity",
    x = "Year"
  ) +
  theme_minimal() +
  theme(axis.title.y.right = element_text(color = "#D84315"))

# Decade comparison
decade_data <- gbif_data %>%
  filter(!is.na(year), year >= 1950) %>%
  mutate(decade = floor(year/10) * 10) %>%
  group_by(decade) %>%
  summarise(
    n_records = n(),
    n_species = n_distinct(species_clean),
    .groups = 'drop'
  ) %>%
  mutate(decade_label = paste0(decade, "s"))

p2 <- ggplot(decade_data, aes(x = decade_label, y = n_species)) +
  geom_bar(stat = "identity", fill = "#1976D2") +
  geom_text(aes(label = n_species), vjust = -0.5) +
  labs(
    title = "Species Recorded per Decade",
    x = "Decade", y = "Number of species"
  ) +
  theme_minimal()

temporal_plot <- p1 / p2
print(temporal_plot)
```

# Elevation Analysis

```{r elevation-analysis, fig.height=10}
# Filter for records with elevation data
elevation_data <- gbif_data %>%
  filter(!is.na(elevation), elevation >= 0, elevation < 5000)  # Remove outliers

if(nrow(elevation_data) > 0) {
  # Overall elevation distribution
  p1 <- ggplot(elevation_data, aes(x = elevation)) +
    geom_histogram(bins = 50, fill = "#5E35B1", alpha = 0.7) +
    geom_density(aes(y = ..count..), color = "#1A237E", size = 1.2) +
    labs(
      title = "Elevation Distribution of Small Mammal Records",
      subtitle = paste("n =", nrow(elevation_data), "records with elevation data"),
      x = "Elevation (m)", y = "Number of records"
    ) +
    theme_minimal()
  
  # Species elevation ranges
  species_elevation <- elevation_data %>%
    group_by(species_clean) %>%
    summarise(
      min_elev = min(elevation),
      max_elev = max(elevation),
      mean_elev = mean(elevation),
      median_elev = median(elevation),
      n_records = n(),
      .groups = 'drop'
    ) %>%
    filter(n_records >= 5) %>%  # Only species with sufficient data
    arrange(median_elev) %>%
    mutate(species_clean = factor(species_clean, levels = species_clean))
  
  p2 <- ggplot(species_elevation, aes(y = species_clean)) +
    geom_segment(aes(x = min_elev, xend = max_elev, yend = species_clean),
                 color = "grey60", size = 2) +
    geom_point(aes(x = median_elev), color = "#E53935", size = 3) +
    geom_point(aes(x = mean_elev), color = "#1976D2", size = 2) +
    labs(
      title = "Elevation Ranges by Species",
      subtitle = "Grey bar: min-max range | Red: median | Blue: mean",
      x = "Elevation (m)", y = ""
    ) +
    theme_minimal() +
    theme(axis.text.y = element_text(face = "italic", size = 9))
  
  elevation_plot <- p1 / p2 + plot_layout(heights = c(1, 2))
  print(elevation_plot)
  
  # Summary statistics
  cat("\nElevation Summary:\n")
  cat("Elevation range:", min(elevation_data$elevation), "-", 
      max(elevation_data$elevation), "m\n")
  cat("Mean elevation:", round(mean(elevation_data$elevation)), "m\n")
  cat("Median elevation:", round(median(elevation_data$elevation)), "m\n")
} else {
  cat("No elevation data available for analysis\n")
}
```

# Data Quality Assessment

## Coordinate Precision Analysis

```{r data-quality, fig.height=8}
# Coordinate precision
precision_summary <- gbif_data %>%
  group_by(coord_precision) %>%
  summarise(
    n_records = n(),
    percent = round(n() / nrow(gbif_data) * 100, 1),
    .groups = 'drop'
  ) %>%
  mutate(coord_precision = factor(coord_precision, 
                                  levels = c("High (<100m)", "Medium (100-1000m)", 
                                           "Low (1-10km)", "Very Low (>10km)", "Unknown")))

p1 <- ggplot(precision_summary, aes(x = coord_precision, y = n_records, fill = coord_precision)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(percent, "%")), vjust = -0.5) +
  scale_fill_manual(values = c("#2E7D32", "#FDD835", "#FF8A65", "#E53935", "grey60")) +
  labs(
    title = "Coordinate Precision of Records",
    subtitle = "Higher precision = more reliable location data",
    x = "Precision Category", y = "Number of records"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

# Basis of record (focal country only, fossils excluded, to match the other panels)
gbif_raw_focal <- gbif_data_raw %>%
  filter(basisOfRecord != "FOSSIL_SPECIMEN",
         datasetKey != "bb5b30b4-827e-4d5e-a86a-825d65cb6583") %>%
  filter(countryCode == focal_iso)

basis_summary <- gbif_raw_focal %>%
  group_by(basisOfRecord) %>%
  summarise(
    n_records = n(),
    percent = round(n() / nrow(gbif_raw_focal) * 100, 1),
    .groups = 'drop'
  ) %>%
  arrange(desc(n_records))

p2 <- ggplot(head(basis_summary, 8), 
             aes(x = reorder(basisOfRecord, n_records), y = n_records)) +
  geom_bar(stat = "identity", fill = "#1976D2") +
  coord_flip() +
  labs(
    title = "Basis of Records",
    subtitle = "Data source types",
    x = "", y = "Number of records"
  ) +
  theme_minimal()

quality_plot <- p1 + p2
print(quality_plot)
```

## Sampling Effort by Province
```{r provincial-effort, fig.height=6}
# Provincial summary
provincial_summary <- gbif_data %>%
  filter(!is.na(stateProvince), stateProvince != "") %>%
  group_by(stateProvince) %>%
  summarise(
    n_records = n(),
    n_species = n_distinct(species_clean),
    n_locations = n_distinct(paste(round(decimalLatitude, 2), 
                                   round(decimalLongitude, 2))),
    .groups = 'drop'
  ) %>%
  arrange(desc(n_records)) %>%
  head(15)  # Top 15 provinces

p_provinces <- ggplot(provincial_summary, 
                      aes(x = reorder(stateProvince, n_records), y = n_records)) +
  geom_bar(stat = "identity", fill = "#00897B") +
  geom_text(aes(label = paste0(n_species, " sp.")), 
            hjust = -0.2, size = 3, color = "#E53935") +
  coord_flip() +
  labs(
    title = "Sampling Effort by Province",
    subtitle = "Bars: number of records | Red text: number of species",
    x = "", y = "Number of records"
  ) +
  theme_minimal()

print(p_provinces)


# Save as PDF instead
ggsave(
  filename = "3_Figures/1_Gbif_overview/Sampling_effort_province_kz.pdf",  # Changed to .pdf
  plot = p_provinces, 
  width = 10,
  height = 8,
  units = "in",
  device = "pdf"  # Explicitly specify PDF
)
```
