---
title: "GBIF Rodent Occurrence Analysis for Central Asia: Kazakhstan Field Guide"
subtitle: "A comprehensive protocol for rodent surveillance planning"
author: "Central Asia One Health Project - ZSL"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: kate
    code_folding: show
    fig_width: 10
    fig_height: 8
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = 'center',
  fig.width = 10,
  fig.height = 8
)

# Load all required libraries
library(tidyverse)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(viridis)
library(scales)
library(patchwork)
library(lubridate)
library(cowplot)
library(kableExtra)
library(plotly)
library(DT)

# Set consistent theme
theme_set(theme_minimal(base_size = 12))

# Define consistent colour palettes
species_colors <- colorRampPalette(c("#440154", "#31688e", "#35b779", "#fde725"))(20)
month_colors <- c("#2E7D32", "#43A047", "#66BB6A", "#81C784", 
                  "#FDD835", "#FFEB3B", "#FFF176", "#FFB74D", 
                  "#FF8A65", "#A1887F", "#8D6E63", "#6D4C41")
family_colors <- c("#E53935", "#D81B60", "#8E24AA", "#5E35B1", 
                   "#3949AB", "#1E88E5", "#039BE5", "#00ACC1", 
                   "#00897B", "#43A047", "#7CB342", "#C0CA33")
```

# Introduction

This document provides a comprehensive analysis of rodent occurrence data from Kazakhstan using GBIF (Global Biodiversity Information Facility) records. It is designed to help field teams prepare for rodent surveillance activities by understanding:

- **What species** you might encounter
- **Where** different species have been recorded
- **When** rodents are most likely to be captured
- **How reliable** the existing data is

> **Note**: This analysis is based on opportunistic occurrence records from GBIF. Patterns may reflect sampling effort and researcher activity rather than true biological patterns. Use this as a guide, not absolute truth.

# Data Loading and Initial Exploration

## Load the GBIF Data
```{r load-data}
# Read the GBIF data
# TO MODIFY: Change this path to your data file location
gbif_data_raw <- read.csv("1_Data_Sets/01_GBIF/Rodents_kazakhstan.csv", 
                          sep = "\t", 
                          stringsAsFactors = FALSE)

# Display basic information
cat("Dataset Overview:\n")
cat("Total records:", format(nrow(gbif_data_raw), big.mark = ","), "\n")
cat("Number of variables:", ncol(gbif_data_raw), "\n")
cat("Date range:", min(gbif_data_raw$year, na.rm = TRUE), "-", 
    max(gbif_data_raw$year, na.rm = TRUE), "\n\n")
```

## Data Cleaning and Preparation
```{r clean-data}
# Create a cleaned dataset with essential fields
gbif_data <- gbif_data_raw %>%
  # Select relevant columns
  dplyr::select(gbifID, scientificName, species, genus, family,
         decimalLatitude, decimalLongitude, 
         coordinateUncertaintyInMeters,
         elevation, year, month, day, eventDate,
         stateProvince, locality,
         basisOfRecord, institutionCode, 
         recordedBy, identifiedBy) %>%
  # Remove records without coordinates
  filter(!is.na(decimalLatitude) & !is.na(decimalLongitude)) %>%
  # Remove records with impossible coordinates
  filter(decimalLatitude >= -90 & decimalLatitude <= 90,
         decimalLongitude >= -180 & decimalLongitude <= 180) %>%
  # Clean species names
  mutate(
    species_clean = ifelse(is.na(species) | species == "", 
                          paste(genus, "sp."), 
                          species),
    family_clean = ifelse(is.na(family) | family == "", 
                          "Unknown", 
                          family),
    # Create date columns
    year = as.integer(year),
    month = as.integer(month),
    # Create coordinate precision categories
    coord_precision = case_when(
      coordinateUncertaintyInMeters <= 100 ~ "High (<100m)",
      coordinateUncertaintyInMeters <= 1000 ~ "Medium (100-1000m)",
      coordinateUncertaintyInMeters <= 10000 ~ "Low (1-10km)",
      coordinateUncertaintyInMeters > 10000 ~ "Very Low (>10km)",
      TRUE ~ "Unknown"
    )
  )

# Summary of data cleaning
cat("Data Cleaning Summary:\n")
cat("Records with coordinates:", format(nrow(gbif_data), big.mark = ","), 
    "(", round(nrow(gbif_data)/nrow(gbif_data_raw)*100, 1), "% retained)\n")
cat("Unique species:", n_distinct(gbif_data$species_clean), "\n")
cat("Unique families:", n_distinct(gbif_data$family_clean), "\n")
cat("Unique genera:", n_distinct(gbif_data$genus), "\n")
```

# Species Diversity Analysis

## Complete Species Inventory

This table shows ALL species recorded in Kazakhstan, sorted by frequency. Every species matters for surveillance planning!
```{r species-inventory}
# Create comprehensive species list
species_summary <- gbif_data %>%
  group_by(species_clean, family_clean) %>%
  summarise(
    n_records = n(),
    n_locations = n_distinct(paste(round(decimalLatitude, 2), 
                                   round(decimalLongitude, 2))),
    year_range = paste(min(year, na.rm = TRUE), "-", 
                      max(year, na.rm = TRUE)),
    n_provinces = n_distinct(stateProvince, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  arrange(desc(n_records)) %>%
  mutate(
    percent_records = round(n_records / sum(n_records) * 100, 2),
    cumulative_percent = round(cumsum(n_records) / sum(n_records) * 100, 1)
  )

# Display interactive table
datatable(species_summary,
          options = list(pageLength = 15, scrollX = TRUE),
          caption = "Complete inventory of rodent species in Kazakhstan GBIF data") %>%
  formatStyle('n_records',
              background = styleColorBar(species_summary$n_records, 'lightblue'),
              backgroundSize = '100% 90%',
              backgroundRepeat = 'no-repeat',
              backgroundPosition = 'center')
```

## Species Accumulation Curve

How complete is our knowledge? This shows if we're still discovering new species.
```{r species-accumulation, fig.height=6}
# Calculate species accumulation by year
species_accumulation <- gbif_data %>%
  filter(!is.na(year)) %>%
  arrange(year) %>%
  group_by(year) %>%
  summarise(
    new_species = list(unique(species_clean)),
    .groups = 'drop'
  ) %>%
  mutate(
    cumulative_species = accumulate(new_species, union),
    n_cumulative = map_int(cumulative_species, length),
    n_new = map_int(new_species, length)
  )

# Plot accumulation curve
p1 <- ggplot(species_accumulation, aes(x = year)) +
  geom_line(aes(y = n_cumulative), size = 1.5, color = "#2E7D32") +
  geom_point(aes(y = n_cumulative), size = 2, color = "#2E7D32") +
  geom_bar(aes(y = n_new * 3), stat = "identity", alpha = 0.3, fill = "#81C784") +
  scale_y_continuous(
    name = "Cumulative number of species",
    sec.axis = sec_axis(~./3, name = "New species per year")
  ) +
  labs(
    title = "Species Discovery Timeline",
    subtitle = "Cumulative species recorded (line) and new species per year (bars)",
    x = "Year"
  ) +
  theme_minimal() +
  theme(
    axis.title.y.right = element_text(color = "#81C784"),
    plot.title = element_text(size = 14, face = "bold")
  )

print(p1)
```

## Family Composition

Understanding which rodent families dominate can help with trap selection and identification training.
```{r family-composition, fig.height=8}
# Prepare family data
family_summary <- gbif_data %>%
  group_by(family_clean) %>%
  summarise(
    n_records = n(),
    n_species = n_distinct(species_clean),
    n_genera = n_distinct(genus),
    .groups = 'drop'
  ) %>%
  arrange(desc(n_records)) %>%
  mutate(
    family_clean = factor(family_clean, levels = family_clean),
    percent = n_records / sum(n_records) * 100
  )

# Create two-panel plot
p1 <- ggplot(family_summary, aes(x = family_clean, y = n_records, fill = family_clean)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = family_colors[1:nrow(family_summary)]) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "A. Number of Records per Family",
    x = "", y = "Number of records"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none",
    plot.title = element_text(face = "bold")
  )

p2 <- ggplot(family_summary, aes(x = family_clean, y = n_species, fill = family_clean)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = family_colors[1:nrow(family_summary)]) +
  labs(
    title = "B. Species Diversity per Family",
    x = "", y = "Number of species"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none",
    plot.title = element_text(face = "bold")
  )

# Combine plots
family_plot <- p1 / p2
print(family_plot)

# Summary statistics
cat("\nFamily Diversity Summary:\n")
cat("Most recorded family:", family_summary$family_clean[1], 
    "with", family_summary$n_records[1], "records\n")
cat("Most species-rich family:", 
    family_summary$family_clean[which.max(family_summary$n_species)],
    "with", max(family_summary$n_species), "species\n")
```

# Spatial Distribution Patterns

## Geographic Coverage Map
```{r coverage-map, fig.height=8}
# Get Kazakhstan boundaries
kazakhstan <- ne_countries(country = "Kazakhstan", scale = "medium", returnclass = "sf")

# Create coverage map with all points
p_coverage <- ggplot() +
  geom_sf(data = kazakhstan, fill = "grey95", color = "grey60") +
  geom_point(data = gbif_data,
             aes(x = decimalLongitude, y = decimalLatitude),
             alpha = 0.5, size = 0.8, color = "#2E7D32") +
  labs(
    title = "Geographic Coverage of Rodent Records in Kazakhstan",
    subtitle = paste("Total:", format(nrow(gbif_data), big.mark = ","), "georeferenced records"),
    x = "Longitude", y = "Latitude"
  ) +
  theme_minimal() +
  coord_sf(xlim = c(46, 88), ylim = c(40, 56))

print(p_coverage)
```

## Species Richness Hotspots

Where is rodent diversity highest? This helps prioritize surveillance locations.
```{r richness-map, fig.height=8}
# Create grid for species richness
grid_size <- 0.5  # degrees

richness_grid <- gbif_data %>%
  mutate(
    lat_grid = floor(decimalLatitude / grid_size) * grid_size + grid_size/2,
    lon_grid = floor(decimalLongitude / grid_size) * grid_size + grid_size/2
  ) %>%
  group_by(lat_grid, lon_grid) %>%
  summarise(
    species_richness = n_distinct(species_clean),
    n_records = n(),
    n_families = n_distinct(family_clean),
    .groups = 'drop'
  )

# Plot species richness
p_richness <- ggplot() +
  geom_sf(data = kazakhstan, fill = "grey95", color = "grey60") +
  geom_tile(data = richness_grid,
            aes(x = lon_grid, y = lat_grid, fill = species_richness),
            alpha = 0.8) +
  scale_fill_viridis(
    name = "Species\nrichness",
    option = "plasma",
    breaks = pretty_breaks(n = 5)
  ) +
  labs(
    title = "Rodent Species Richness Hotspots",
    subtitle = paste("Grid size:", grid_size, "degrees"),
    x = "Longitude", y = "Latitude"
  ) +
  theme_minimal() +
  coord_sf(xlim = c(46, 88), ylim = c(40, 56))

print(p_richness)

# Identify top diversity hotspots
top_hotspots <- richness_grid %>%
  arrange(desc(species_richness)) %>%
  head(10) %>%
  mutate(
    location = paste0("(", round(lat_grid, 1), "°N, ", round(lon_grid, 1), "°E)")
  )

cat("\nTop 10 Diversity Hotspots:\n")
for(i in 1:nrow(top_hotspots)) {
  cat(i, ". ", top_hotspots$location[i], ": ", 
      top_hotspots$species_richness[i], " species (", 
      top_hotspots$n_records[i], " records)\n", sep = "")
}
```

## Individual Species Distribution Maps

Interactive maps for the most common species AND rare species (important for surveillance!).
```{r species-maps, fig.height=12}
# Select species to map: top 6 common + 3 rare species with >1 record
common_species <- species_summary %>%
  head(6) %>%
  pull(species_clean)

rare_species <- species_summary %>%
  filter(n_records > 1) %>%
  tail(3) %>%
  pull(species_clean)

species_to_map <- c(common_species, rare_species)

# Create individual maps
species_maps <- map(species_to_map, function(sp) {
  sp_data <- gbif_data %>% filter(species_clean == sp)
  
  p <- ggplot() +
    geom_sf(data = kazakhstan, fill = "grey95", color = "grey60") +
    geom_point(data = sp_data,
               aes(x = decimalLongitude, y = decimalLatitude),
               color = "#E53935", alpha = 0.7, size = 1.5) +
    labs(
      title = sp,
      subtitle = paste("n =", nrow(sp_data), "records")
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(face = "italic", size = 10),
      plot.subtitle = element_text(size = 8)
    ) +
    coord_sf(xlim = c(46, 88), ylim = c(40, 56))
  
  return(p)
})

# Arrange in grid
species_grid <- wrap_plots(species_maps, ncol = 3) +
  plot_annotation(
    title = "Distribution Maps: Common Species (top 2 rows) and Rare Species (bottom row)",
    theme = theme(plot.title = element_text(size = 14, face = "bold"))
  )

print(species_grid)
```

# Temporal Patterns

## Seasonal Activity Patterns

**Critical Note**: Seasonal patterns may reflect researcher activity (holidays, field seasons) rather than true rodent activity!
```{r seasonal-patterns, fig.height=10}
# Prepare monthly data
monthly_data <- gbif_data %>%
  filter(!is.na(month)) %>%
  group_by(month) %>%
  summarise(
    n_records = n(),
    n_species = n_distinct(species_clean),
    .groups = 'drop'
  ) %>%
  mutate(
    month_name = month.abb[month],
    month_name = factor(month_name, levels = month.abb)
  )

# Overall monthly pattern
p1 <- ggplot(monthly_data, aes(x = month_name, y = n_records)) +
  geom_bar(stat = "identity", fill = month_colors) +
  geom_line(aes(group = 1, y = n_species * max(n_records) / max(n_species)), 
            color = "red", size = 1.5) +
  scale_y_continuous(
    name = "Number of records",
    labels = comma,
    sec.axis = sec_axis(~. * max(monthly_data$n_species) / max(monthly_data$n_records), 
                        name = "Number of species (red line)")
  ) +
  labs(
    title = "A. Overall Seasonal Pattern",
    subtitle = "⚠️ May reflect sampling effort rather than biological patterns",
    x = "Month"
  ) +
  theme_minimal() +
  theme(
    axis.title.y.right = element_text(color = "red"),
    plot.subtitle = element_text(color = "darkorange", face = "italic")
  )

# Family-specific patterns
family_monthly <- gbif_data %>%
  filter(!is.na(month), family_clean %in% head(family_summary$family_clean, 6)) %>%
  group_by(family_clean, month) %>%
  summarise(n_records = n(), .groups = 'drop') %>%
  mutate(month_name = factor(month.abb[month], levels = month.abb))

p2 <- ggplot(family_monthly, aes(x = month_name, y = n_records, fill = family_clean)) +
  geom_bar(stat = "identity") +
  facet_wrap(~family_clean, scales = "free_y", ncol = 2) +
  scale_fill_manual(values = family_colors[1:6]) +
  labs(
    title = "B. Seasonal Patterns by Family",
    x = "Month", y = "Number of records"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8)
  )

seasonal_plot <- p1 / p2 + plot_layout(heights = c(1, 2))
print(seasonal_plot)

# Statistical summary
cat("\nSeasonal Sampling Summary:\n")
cat("Peak collection month:", monthly_data$month_name[which.max(monthly_data$n_records)], 
    "with", max(monthly_data$n_records), "records\n")
cat("Lowest collection month:", monthly_data$month_name[which.min(monthly_data$n_records)], 
    "with", min(monthly_data$n_records), "records\n")
cat("Months with no records:", paste(month.abb[!(1:12 %in% monthly_data$month)], collapse = ", "), "\n")
```

## Long-term Temporal Trends
```{r temporal-trends, fig.height=8}
# Yearly patterns
yearly_data <- gbif_data %>%
  filter(!is.na(year), year >= 1950) %>%  # Focus on modern records
  group_by(year) %>%
  summarise(
    n_records = n(),
    n_species = n_distinct(species_clean),
    n_collectors = n_distinct(recordedBy, na.rm = TRUE),
    .groups = 'drop'
  )

p1 <- ggplot(yearly_data, aes(x = year)) +
  geom_bar(aes(y = n_records), stat = "identity", fill = "#2E7D32", alpha = 0.7) +
  geom_line(aes(y = n_species * 10), color = "#D84315", size = 1.2) +
  scale_y_continuous(
    name = "Number of records",
    labels = comma,
    sec.axis = sec_axis(~./10, name = "Number of species")
  ) +
  labs(
    title = "Collection Effort Over Time",
    subtitle = "Bars: records per year | Line: species diversity",
    x = "Year"
  ) +
  theme_minimal() +
  theme(axis.title.y.right = element_text(color = "#D84315"))

# Decade comparison
decade_data <- gbif_data %>%
  filter(!is.na(year), year >= 1950) %>%
  mutate(decade = floor(year/10) * 10) %>%
  group_by(decade) %>%
  summarise(
    n_records = n(),
    n_species = n_distinct(species_clean),
    .groups = 'drop'
  ) %>%
  mutate(decade_label = paste0(decade, "s"))

p2 <- ggplot(decade_data, aes(x = decade_label, y = n_species)) +
  geom_bar(stat = "identity", fill = "#1976D2") +
  geom_text(aes(label = n_species), vjust = -0.5) +
  labs(
    title = "Species Recorded per Decade",
    x = "Decade", y = "Number of species"
  ) +
  theme_minimal()

temporal_plot <- p1 / p2
print(temporal_plot)
```

# Elevation Analysis

Understanding elevation preferences helps with site selection in mountainous regions.
```{r elevation-analysis, fig.height=10}
# Filter for records with elevation data
elevation_data <- gbif_data %>%
  filter(!is.na(elevation), elevation >= 0, elevation < 5000)  # Remove outliers

if(nrow(elevation_data) > 0) {
  # Overall elevation distribution
  p1 <- ggplot(elevation_data, aes(x = elevation)) +
    geom_histogram(bins = 50, fill = "#5E35B1", alpha = 0.7) +
    geom_density(aes(y = ..count..), color = "#1A237E", size = 1.2) +
    labs(
      title = "Elevation Distribution of Rodent Records",
      subtitle = paste("n =", nrow(elevation_data), "records with elevation data"),
      x = "Elevation (m)", y = "Number of records"
    ) +
    theme_minimal()
  
  # Species elevation ranges
  species_elevation <- elevation_data %>%
    group_by(species_clean) %>%
    summarise(
      min_elev = min(elevation),
      max_elev = max(elevation),
      mean_elev = mean(elevation),
      median_elev = median(elevation),
      n_records = n(),
      .groups = 'drop'
    ) %>%
    filter(n_records >= 5) %>%  # Only species with sufficient data
    arrange(median_elev) %>%
    mutate(species_clean = factor(species_clean, levels = species_clean))
  
  p2 <- ggplot(species_elevation, aes(y = species_clean)) +
    geom_segment(aes(x = min_elev, xend = max_elev, yend = species_clean),
                 color = "grey60", size = 2) +
    geom_point(aes(x = median_elev), color = "#E53935", size = 3) +
    geom_point(aes(x = mean_elev), color = "#1976D2", size = 2) +
    labs(
      title = "Elevation Ranges by Species",
      subtitle = "Grey bar: min-max range | Red: median | Blue: mean",
      x = "Elevation (m)", y = ""
    ) +
    theme_minimal() +
    theme(axis.text.y = element_text(face = "italic", size = 9))
  
  elevation_plot <- p1 / p2 + plot_layout(heights = c(1, 2))
  print(elevation_plot)
  
  # Summary statistics
  cat("\nElevation Summary:\n")
  cat("Elevation range:", min(elevation_data$elevation), "-", 
      max(elevation_data$elevation), "m\n")
  cat("Mean elevation:", round(mean(elevation_data$elevation)), "m\n")
  cat("Median elevation:", round(median(elevation_data$elevation)), "m\n")
} else {
  cat("No elevation data available for analysis\n")
}
```

# Data Quality Assessment

## Coordinate Precision Analysis

Understanding data quality helps assess reliability of distribution patterns.
```{r data-quality, fig.height=8}
# Coordinate precision
precision_summary <- gbif_data %>%
  group_by(coord_precision) %>%
  summarise(
    n_records = n(),
    percent = round(n() / nrow(gbif_data) * 100, 1),
    .groups = 'drop'
  ) %>%
  mutate(coord_precision = factor(coord_precision, 
                                  levels = c("High (<100m)", "Medium (100-1000m)", 
                                           "Low (1-10km)", "Very Low (>10km)", "Unknown")))

p1 <- ggplot(precision_summary, aes(x = coord_precision, y = n_records, fill = coord_precision)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(percent, "%")), vjust = -0.5) +
  scale_fill_manual(values = c("#2E7D32", "#FDD835", "#FF8A65", "#E53935", "grey60")) +
  labs(
    title = "Coordinate Precision of Records",
    subtitle = "Higher precision = more reliable location data",
    x = "Precision Category", y = "Number of records"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

# Basis of record
basis_summary <- gbif_data_raw %>%
  group_by(basisOfRecord) %>%
  summarise(
    n_records = n(),
    percent = round(n() / nrow(gbif_data_raw) * 100, 1),
    .groups = 'drop'
  ) %>%
  arrange(desc(n_records))

p2 <- ggplot(head(basis_summary, 8), 
             aes(x = reorder(basisOfRecord, n_records), y = n_records)) +
  geom_bar(stat = "identity", fill = "#1976D2") +
  coord_flip() +
  labs(
    title = "Basis of Records",
    subtitle = "Data source types",
    x = "", y = "Number of records"
  ) +
  theme_minimal()

quality_plot <- p1 + p2
print(quality_plot)
```

## Sampling Effort by Province
```{r provincial-effort, fig.height=6}
# Provincial summary
provincial_summary <- gbif_data %>%
  filter(!is.na(stateProvince), stateProvince != "") %>%
  group_by(stateProvince) %>%
  summarise(
    n_records = n(),
    n_species = n_distinct(species_clean),
    n_locations = n_distinct(paste(round(decimalLatitude, 2), 
                                   round(decimalLongitude, 2))),
    .groups = 'drop'
  ) %>%
  arrange(desc(n_records)) %>%
  head(15)  # Top 15 provinces

p_provinces <- ggplot(provincial_summary, 
                      aes(x = reorder(stateProvince, n_records), y = n_records)) +
  geom_bar(stat = "identity", fill = "#00897B") +
  geom_text(aes(label = paste0(n_species, " sp.")), 
            hjust = -0.2, size = 3, color = "#E53935") +
  coord_flip() +
  labs(
    title = "Sampling Effort by Province",
    subtitle = "Bars: number of records | Red text: number of species",
    x = "", y = "Number of records"
  ) +
  theme_minimal()

print(p_provinces)
```

# Field Planning Recommendations

Based on this analysis, here are key recommendations for field teams:

## Spatial Priorities
```{r recommendations-spatial}
# Identify undersampled areas
undersampled <- richness_grid %>%
  filter(n_records < 10, species_richness > 0) %>%
  arrange(desc(species_richness)) %>%
  head(5) %>%
  mutate(location = paste0(round(lat_grid, 1), "°N, ", round(lon_grid, 1), "°E"))

cat("UNDERSAMPLED AREAS WITH KNOWN DIVERSITY:\n")
cat("These locations have species records but low sampling effort:\n\n")
for(i in 1:nrow(undersampled)) {
  cat("• ", undersampled$location[i], ": ", 
      undersampled$species_richness[i], " species from only ", 
      undersampled$n_records[i], " records\n", sep = "")
}

cat("\n")
cat("HIGH PRIORITY SPECIES (rare but important):\n")
rare_important <- species_summary %>%
  filter(n_records <= 10, n_records > 1) %>%
  head(10)
for(i in 1:nrow(rare_important)) {
  cat("• ", rare_important$species_clean[i], ": only ", 
      rare_important$n_records[i], " records from ", 
      rare_important$n_locations[i], " location(s)\n", sep = "")
}
```

## Temporal Recommendations
```{r recommendations-temporal}
# Best months for sampling
best_months <- monthly_data %>%
  arrange(desc(n_species)) %>%
  head(3)

cat("OPTIMAL SAMPLING MONTHS (based on historical species diversity):\n")
for(i in 1:nrow(best_months)) {
  cat(i, ". ", best_months$month_name[i], ": ", 
      best_months$n_species[i], " species recorded\n", sep = "")
}

cat("\n⚠️ Remember: These patterns may reflect researcher availability!\n")
cat("Consider year-round monitoring to capture true seasonal patterns.\n")
```

## Species-Specific Considerations
```{r species-considerations}
# Create a summary table for field identification
field_guide <- species_summary %>%
  head(20) %>%
  dplyr::select(Species = species_clean, 
         Family = family_clean,
         `Records` = n_records,
         `Locations` = n_locations,
         `Provinces` = n_provinces) %>%
  mutate(Species = cell_spec(Species, italic = TRUE))

kable(field_guide, escape = FALSE, caption = "Top 20 Species Quick Reference") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = FALSE) %>%
  column_spec(1, bold = TRUE) %>%
  row_spec(1:5, background = "#E8F5E9")
```

# Data Export for Field Use
```{r export-data}
# Create field-ready datasets
# TO MODIFY: Change export paths as needed

# 1. Species checklist
species_checklist <- species_summary %>%
  dplyr::select(species = species_clean, 
         family = family_clean, 
         n_records, 
         n_locations,
         year_range)

write.csv(species_checklist, 
          "kazakhstan_rodent_checklist.csv", 
          row.names = FALSE)

# 2. Hotspot locations
hotspot_export <- richness_grid %>%
  arrange(desc(species_richness)) %>%
  head(50) %>%
  mutate(
    latitude = lat_grid,
    longitude = lon_grid
  ) %>%
  dplyr::select(latitude, longitude, species_richness, n_records, n_families)

write.csv(hotspot_export, 
          "kazakhstan_rodent_hotspots.csv", 
          row.names = FALSE)

# 3. Monthly activity calendar
monthly_export <- monthly_data %>%
  dplyr::select(month, month_name, n_records, n_species)

write.csv(monthly_export, 
          "kazakhstan_rodent_seasonal.csv", 
          row.names = FALSE)

cat("✅ Data exported to 'outputs' folder:\n")
cat("• kazakhstan_rodent_checklist.csv\n")
cat("• kazakhstan_rodent_hotspots.csv\n")
cat("• kazakhstan_rodent_seasonal.csv\n")
```

# Session Information
```{r session-info}
sessionInfo()
```

---

*Document generated on `r Sys.Date()` for the Central Asia One Health Project*

*Remember: GBIF data represents opportunistic sampling. Absence of records doesn't mean absence of species!*